---
title: "Models for Exploratory Data Analysis - Linear Regression"
date: "Nov-2019"
output: 
  rmarkdown::github_document:
    fig_width: 9
    fig_height: 5
    dev: png
---

```{r setup, include=FALSE}
options(scipen = 9999)
```

This analysis applies linear regression to data on house prices in England and Wales. It is intended to demonstrate that fitting simple models to a dataset can be a highly effective way of quickly identifying trends and generating ideas for further analysis. 

More information on this approach can be found here: https://r4ds.had.co.nz/model-building.html 

## How have house prices in England and Wales changed since 2008? 

Data on median property prices in England and Wales is published quarterly by the UK's Office for National Statistics, using data from the Land Registry. It is available here: https://www.ons.gov.uk/peoplepopulationandcommunity/housing/datasets/medianhousepricefornationalandsubnationalgeographiesquarterlyrollingyearhpssadataset09   

The Excel document contains house prices for different administrative geographies of England and Wales, and different types of property. In this analysis, we will look at house prices in the final quarter of each year since 2008 across c. 350 local authority areas in England and Wales. 

We'll start by reading in the data and do some basic cleaning. 

```{r, message=FALSE}
library(tidyverse)
library(readxl)
library(lubridate)

house_prices <- read_xls("hpssadataset9medianpricepaidforadministrativegeographies.xls",
         sheet = "2a", skip = 6) %>% 
  # Converting the data from 'wide' to long
  gather(key = "date", value = "median_price", -1, -2, -3, -4) %>% 
  # Converting date to a date format using the myd() function from the lubridate package
  filter(str_detect(date, "Dec")) %>% 
  mutate(year = readr::parse_number(date)) %>% 
  # Looking at data since 2008
  filter(year >= 2008) %>% 
  janitor::clean_names() %>% 
  select(-region_country_code, -date)

# Clean data 
house_prices 

```

## Visualising changes to house prices

Let’s quickly graph the changes to house prices.

```{r, message=FALSE}
library(scales)

house_prices %>% 
  ggplot(aes(year, median_price)) +
  geom_line(aes(group = local_authority_name), alpha = 0.2) + 
  geom_smooth() +
  scale_y_continuous(labels = dollar_format(prefix = "£")) + 
  labs(x = "", y = "Median house prices")

house_prices %>% 
  ggplot(aes(year, median_price)) +
  geom_line(aes(group = local_authority_name), alpha = 0.2) + 
  geom_smooth() + 
  scale_y_continuous(trans = "log10", labels = dollar_format(prefix = "£")) + 
  annotation_logticks(side = "l") +
  labs(x = "", y = "Median house prices (log scale)")
  
```

 At first glance it appears that house prices have been steadily rising across England and Wales, though at very different rates. There is too much data (even in this relatively small dataset) to easily identify any key takeaways - so we need to find a way of summarising the data.   

## Fitting a (simple) model 

One way of summarising the data is to fit a model that captures the general trend. Since what we're interested in is how house prices have changed since 2008, we should try to fit a model that can capture the variation over time. It would also be interesting to look at whether local authorities that are close together have experienced similar trends in house price changes. 

It's generally better during an exploratory exercise to start with a simple model, and then expand. The first thing we'll do is visualise what a simple linear model would look like when applied to one local authority - here, we'll used Ealing in West London. This model only uses the passage of time to predict house prices - it is not a sophisticated model, but merely captures the trend in how house prices have changed over time.      

```{r}

house_prices %>% 
  filter(local_authority_name == "Ealing") %>% 
  ggplot(aes(year, median_price)) +
  geom_line() + 
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) + 
  scale_y_continuous(labels = dollar_format(prefix = "£")) + 
  labs(y = "Median Price")

```

This model does a reasonable job of summarising the overall trend in house prices in Ealing, with the blue fitted line generally close to the black line showing house price data. Now let's see how well this approach scales to c. 350 local authorities. 

## Fitting 348 (simple) models

We'll use the nested dataframe workflow to split our data into separate datasets - one for each local authority. We'll then fit the same linear regression model to each one, and calculate a couple of summary measures to compare the quality and findings of the model. 

Note that for this model we'll log transform the median prices to base 2. This flattens out some of the variation in the prices data, and means our resulting coefficients can (after some transformation) be interpreted as the estimated annual percentage change in house prices. 

```{r, message=FALSE}
library(modelr)
library(broom)
library(scales)

house_prices_model <- house_prices %>% 
  mutate(year_since_08 = year-2008) %>% 
  group_by(region_country_name, local_authority_name, local_authority_name) %>% 
  nest() %>%  
  ungroup() %>% 
  mutate(price_model = map(data, 
                           ~ lm(log2(.$median_price) ~ .$year_since_08, data = .))) %>% 
  mutate(model_summary = map(price_model, glance), 
         tidy_model = map(price_model, tidy)) 

```

## How good are the models? 

We'll use R squared to assess how good our models are. Broadly, an R squared of 1 indicates a perfect fit, while 0 indicates a very poor fit and a model with no explanatory power. 

```{r}

model_summary <- house_prices_model %>% 
  unnest(model_summary) %>% 
  select(-data, -price_model) %>% 
  janitor::clean_names() 

model_summary %>% 
  ggplot(aes(fct_reorder(region_country_name, r_squared), 
             r_squared, col = region_country_name)) +
  geom_boxplot(outlier.shape = NULL, varwidth = TRUE) + 
  geom_jitter(alpha = 0.5) + 
  coord_flip() +
  theme(legend.position = "none") + 
  labs(x = "", y = "R Square")

```

Here we can see that locations in London, the South East and the East have the highest R squared, which means the models work well here. The fit is likely to look like the graph for Ealing above. 

We can visualise what higher levels of R square indicate in terms of trends in the data. Below, we've 'cut' r squared into 5 equally-spaced groups. We can see that areas with higher levels of R square have seen faster rates of house price growth. In locations where R squared is very low, house prices have barely moved at all or have been very volatile.   

```{r message=FALSE} 

house_prices %>% 
  inner_join(model_summary) %>% 
  mutate(r_squared_cut = cut_width(r_squared, 0.2, boundary = 0)) %>% 
  ggplot(aes(year, median_price)) + 
  geom_line(aes(col = r_squared_cut, group = local_authority_name), alpha = 0.5, show.legend = FALSE) + 
  geom_smooth() + 
  facet_wrap(~r_squared_cut, nrow = 1) +
  scale_y_continuous(trans = "log10", labels = dollar_format(prefix = "£")) +
  scale_x_continuous(breaks = seq(2008, 2018, 2)) + 
  labs(x = "", y = "Median Price (£)")

```

## Interpreting the results

We can transform the results of the models to assess how much house prices have changed since 2008 in different parts of England and Wales. The transformed coefficients of our model (see code below) can be interpreted as estimated annual % change in house prices since 2008. 

```{r, message = TRUE}

model_tidy <- house_prices_model %>% 
  unnest(tidy_model, .drop = TRUE) %>% 
  select(-std.error, -statistic, -p.value) %>% 
  spread(term, value = estimate) %>% 
  janitor::clean_names() %>% 
  mutate_at(vars("year_since_08","intercept"), ~ 2^.) %>% 
  mutate(year_since_08 = year_since_08-1) %>% 
  inner_join(model_summary)

model_tidy %>% 
  arrange(intercept) %>% 
  ggplot(aes(r_squared, year_since_08, fill = region_country_name)) + 
  geom_point(aes(size = intercept), shape = 21, alpha = 0.5) + 
  scale_y_continuous(labels = percent) + 
  labs(x = "R square", y = "Estimated annual increases in house prices (%)")

```

Here, we can see that locations in London, the East and the East Midlands have seen the biggest increases to house prices. The model fit in these locations, as measured by R squared, is also very high. 

We can visualise what this means in practise by adding the predictions back into the data. As we can see below, the predicted values for Tower Hamlets (a borough in East London) closely track the median house prices. 

```{r}

model_tidy %>% 
  filter(local_authority_name == "Tower Hamlets") %>% 
  select(local_authority_name, data, price_model) %>% 
  mutate(predictions = map2(data, price_model, add_predictions, var = "price_pred")) %>% 
  unnest(predictions) %>% 
  mutate(price_pred = 2 ^ price_pred) %>% 
  ggplot(aes(year, price_pred)) +
  geom_line(col = "blue", linetype = "dashed") + 
  geom_point(aes(y = median_price)) +
  scale_y_continuous(labels = dollar_format(prefix = "£")) +
  scale_x_continuous(breaks = seq(2008, 2018, 2)) + 
  labs(x = "", y = "", 
       title = "Actual and predicted median house prices in Tower Hamlets, 2008 - 2018")

```

We can dig deeper into the variation in model performance by region by faceting our earlier visualisation. Below, we can see that almost everywhere in London has experienced more than 6% annual increases in house prices since 2008, while in the East and South East more than half of locations have experienced more than 5% annual increases. Most locations in the East Midlands, South West and West Midlands have experienced between 2.5% and 5%. In Yorkshire, the North West, North East and Wales, the coefficients are lower and model performance worse - likely indicating that house price growth has been slower or more volatile. 

```{r}

model_tidy %>% 
  mutate(region_country_name = 
           fct_reorder(region_country_name, -year_since_08, .fun = median)) %>% 
  ggplot(aes(r_squared, year_since_08, fill = region_country_name)) + 
  geom_point(shape = 21, alpha = 0.5) + 
  scale_y_continuous(labels = percent) + 
  facet_wrap(~region_country_name, nrow = 2) + 
  labs(x = "R square", y = "Estimated annual increases in house prices (%)") + 
  theme(legend.position = "none")

```

We can examine the differences in performance within Wales if we pull-out some of the 'best' and 'worst' fitting locations. While the best fitting locations have undoubtedly seen slower growth than those in London and the South East, they have seen some growth. However, areas where the model fits badly have seen almost no growth in house prices over the last decade.   

```{r, message=FALSE}

wales_best <- model_tidy %>% 
  filter(region_country_name == "Wales") %>% 
  arrange(desc(r_squared)) %>% 
  head(5)

wales_worst <- model_tidy %>% 
  filter(region_country_name == "Wales") %>% 
  arrange(desc(r_squared)) %>% 
  tail(5)

wales_modlels <- bind_rows(list("best" = wales_best, "worst" = wales_worst), .id = "fit_quality")

house_prices %>% 
  inner_join(wales_modlels) %>% 
  ggplot(aes(year, median_price)) + 
  geom_line(aes(group = local_authority_name, col = fit_quality), show.legend = FALSE) + 
  geom_smooth() +
  facet_wrap(~fit_quality) +
  scale_y_continuous(labels = dollar_format(prefix = "£")) + 
  labs(x = "", y = "Median House Prices", 
       title = "Median House Prices (2008 vs 2018) in Best and Worst Fitting Areas in Wales") + 
  theme(legend.position = "none")

house_prices %>% 
  inner_join(wales_modlels) %>% 
  filter(year %in% c(2008, 2018)) %>% 
  ggplot(aes(factor(year), median_price, col = fit_quality)) + 
  geom_line(aes(group = local_authority_name)) + 
  geom_point() +
  geom_text(data = house_prices %>% 
              inner_join(wales_modlels) %>% 
              filter(year == 2018), 
            aes(label = local_authority_name), hjust = "left") + 
  facet_wrap(~fit_quality) + 
  scale_y_continuous(labels = dollar_format(prefix = "£")) + 
  labs(x = "", y = "Median House Prices", 
       title = "Median House Prices (2008 vs 2018) in Best and Worst Fitting Areas in Wales") + 
  theme(legend.position = "none")

```
